|

### 2023-05-14 paper/00-main.Rmd
more edits/writing

abstract: |
  In American football, a pass rush is an attempt by the defensive team to disrupt the offense and prevent the quarterback (QB) from completing a pass. Existing metrics for assessing pass rush performance are either discrete-time quantities or based on subjective judgement. Using player tracking data, we propose STRAIN, a novel metric for evaluating pass rushers in the National Football League (NFL) at the continuous-time within-play level. Inspired by the concept of strain rate in materials science, STRAIN is a simple and interpretable means for measuring defensive pressure in football. It is a directly-observed statistic as a function of two features: the distance between the pass rusher and QB, and the rate at which this distance is being reduced. Our metric possesses great predictability of pressure and stability over time. We use a multilevel model to isolate the contribution of every pass rusher in a play to the average STRAIN while controlling for their pass blocker matchup, the QB, and play-level information. We apply our approach to NFL data and present results for the first 8 weeks of the 2021 season. In particular, we provide comparisons of STRAIN for different defensive positions and play outcomes, and rankings of the NFL's best pass rushers according to our metric.

@@

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
  # fig.pos = "p",
  # cache = TRUE
)
library(here)
library(tidyverse)
```

@@

All code related to this paper is available at https://github.com/getstrained/intro-strain. The data provided by the NFL Big Data Bowl 2023 is available at: https://www.kaggle.com/competitions/nfl-big-data-bowl-2023/data.

>>>>

abstract: |
  In American football, a pass rush is an attempt by the defensive team to disrupt the offense and prevent the quarterback (QB) from completing a pass. Existing metrics for assessing pass rush performance are either discrete-time quantities or based on subjective judgement. Using player tracking data, we propose STRAIN, a novel metric for evaluating pass rushers in the National Football League (NFL) at the continuous-time within-play level. Inspired by the concept of strain rate in materials science, STRAIN is a simple and interpretable means for measuring defensive pressure in football. It is a directly-observed statistic as a function of two features: the distance between the pass rusher and QB, and the rate at which this distance is being reduced. Our metric possesses great predictability of pressure and stability over time. We also fit a multilevel model for STRAIN to understand the defensive pressure contribution of every pass rusher at the play-level. We apply our approach to NFL data and present results for the first 8 weeks of the 2021 season. In particular, we provide comparisons of STRAIN for different defensive positions and play outcomes, and rankings of the NFL's best pass rushers according to our metric.

@@

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  # fig.pos = "p",
  cache = TRUE
)
library(here)
library(tidyverse)
```

@@

All code related to this paper is available at https://github.com/getstrained/intro-strain. The data provided by the NFL Big Data Bowl 2023 is available at https://www.kaggle.com/competitions/nfl-big-data-bowl-2023/data.

### 2023-05-14 paper/01-introduction.Rmd
more edits/writing

In an attempt to foster analytics and innovate the game, the National Football League (NFL) introduced their player tracking system known as [Next Gen Stats (NGS)](https://operations.nfl.com/gameday/technology/nfl-next-gen-stats/) in 2016, using radio frequency identification (RFID) chips placed in players' shoulder pads (and in the ball) to collect data 10 times per second. These NGS tracking data contain information about the on-field locations, speeds, and angles of all 22 players on the field (as well as the ball). While these data were initially only available for teams, media, and vendors, in December 2018 the NFL launched the inaugural edition of their annual [Big Data Bowl (BDB)](https://operations.nfl.com/gameday/analytics/big-data-bowl/past-big-data-bowl-recaps/) competition.

The first BDB led to several contributions largely focused on offensive performance evaluation. For example, one group of finalists introduced an approach for modeling the hypothetical completion probability of a pass aiding in the evaluation of quarterback (QB) decision making [@deshpande2020expected]. The winners of the inaugural BDB focused on identifying receiver routes via clustering techniques [@chu2020route] and convolutional neural networks [@sterken2019routenet]. Along with the BDB competition entries, the public release of NGS data allowed researchers to tackle (pun intended) a variety of other problems such as revisiting fourth down decision making [@lopez2020bigger], annotating pass coverage with Gaussian mixture models [@dutta2020unsupervised], and introducing a continuous-time framework to estimate within-play value [@yurko2020going].  Since its inception, the BDB has chosen a different theme each year leading to new insight about evaluating different positions such as running backs, defensive backs, and special teams. The 2023 edition of the NFL Big Data Bowl asked participants to evaluate linemen on passing plays [@Howard2023NFL].

We limit our attention in this manuscript to the defensive side of the ball. There are two main types of defensive linemen in American football: (1) defensive tackles (DTs) located within the interior of the line and (2) defensive ends (DEs) located along the edges (see Figure \ref{fig:formation} (top) for example formation with DTs and DEs). The primary purpose of both positions is to rush the QB on passing plays, with DEs displaying superiority in observed pass rushing ability [@Eager2018nfl]. Additionally, within DTs there are nose tackles (NTs) who directly line up across from the ball at the line of scrimmage (see Figure \ref{fig:formation} (bottom) for example defensive formation with a NT). Typically, teams will employ either one NT or two DTs on the interior with DEs along either side of the defensive line. In addition to defensive lineman, other positions may attempt to rush the QB on blitzing plays such as outside linebackers (OLBs), interior linebackers (ILBs), and potentially members of the secondary (cornerback, free safety, strong safety) who typically play pass coverage.

@@

Moreover, the sum of sacks, hits, and hurries is often defined as (QB) pressures.
This is somewhat better than sacks, but suffers from problems of subjectivity (e.g., whether there is an actual hurry or not).
Pro football focus (PFF) comes up with a metric called "pass-rush productivity" (PRP), which is a minor modification from the aforementioned pressures metric (see Table \ref{tab:metrics}).
In particular, PRP gives twice as much weight to a sack relative to hurries and hits, which is a small upgrade to pressures.
However, the choice of weights is ad-hoc and still only measures binary outcomes, similar to the shortcomings of previously discussed metrics.

@@

More recently proposed metrics such as time in pocket (TIP), NGS Get Off [@Hermsmeyer2021nfl], and pass rush win rate (PRWR) [@Burke2018Created] are substantial improvements over the less sophisticated counting statistics, but nevertheless are still imperfect.
TIP refers to how long a QB can operate within the protected space behind the offensive line, known as the pocket.
However, this measure is highly context-dependent, as it can be influenced by a number of factors such as the defensive scheme or type of passing route.
NGS Get Off is an aggregated statistic, illustrating how quickly a defender can get past the line of scrimmage after the snap on average.
PRWR is created using player tracking data, which is at a more granular level than previous measures.
It demonstrates whether a pass rusher is able to beat their blocking matchup before a fixed time from the snap (2.5 seconds as chosen by ESPN).
This depends on the rather arbitrary time threshold used to define a pass rush win.
In addition, once a cutoff is chosen, this metric converts continuous data to a win-loss indicator, becoming binary like most of the metrics discussed above.

@@

The peer-reviewed literature on measuring the performance of football linemen (either offensive or defensive) is scant.
@AlamarGould2008 find an association between pass completion rate and successful pass blocking by offensive linemen.
The data for this study are collected for the first three weeks of the 2007 NFL season, manually recording whether a lineman holds a block and the time it took for quarterback to throw the football.
@AlamarGoldner2011 later follow up by using manually-tracked data for the 2010 season to estimate lineman performance for different team-positions instead of individual defenders (e.g. Chicago Bears' left tackle, Pittsburgh Steelers' center).
The paper uses survival analysis to model time in pocket and completion percentage for quarterbacks, before proposing a measure for linemen's contribution to their team's passing in terms of yards gained.

>>>>

In an attempt to foster analytics and innovate the game, the National Football League (NFL) introduced their player tracking system known as Next Gen Stats (NGS) in 2016 [@nfl2023ngs].
NGS uses radio frequency identification (RFID) chips placed in players' shoulder pads (and in the ball) to collect data at a rate of 10 frames per second.
The data captures real-time on-field information such as locations, speeds, and accelerations of all 22 players (as well as the ball).
While these data were initially only available for teams, media, and vendors, in December 2018 the NFL launched the inaugural edition of their annual Big Data Bowl (BDB) competition [@nfl2023big].

The first BDB led to several contributions largely focused on offensive performance evaluation. For example, one group of finalists introduced an approach for modeling the hypothetical completion probability of a pass aiding in the evaluation of quarterback (QB) decision making [@deshpande2020expected]. The winners of the inaugural BDB focused on identifying receiver routes via clustering techniques [@chu2020route] and convolutional neural networks [@sterken2019routenet]. Along with the BDB competition entries, the public release of NGS data allowed researchers to tackle <!-- (pun intended) --> a variety of other problems such as revisiting fourth down decision making [@lopez2020bigger], annotating pass coverage with Gaussian mixture models [@dutta2020unsupervised], and introducing a continuous-time framework to estimate within-play value [@yurko2020going].  Since its inception, the BDB has chosen a different theme each year leading to new insight about evaluating different positions such as running backs, defensive backs, and special teams. The 2023 edition of the NFL Big Data Bowl asked participants to evaluate linemen on passing plays [@Howard2023NFL].

In this manuscript, we specifically focus on measuring the performance of defensive linemen in the NFL. There are two main types of defensive linemen in American football: (1) defensive tackles (DTs) located within the interior of the line and (2) defensive ends (DEs) located along the edges (see Figure \ref{fig:formation} (top) for example formation with DTs and DEs). The primary purpose of both positions is to rush the QB on passing plays, with DEs displaying superiority in observed pass rushing ability [@Eager2018nfl]. Additionally, within DTs there are nose tackles (NTs) who directly line up across from the ball at the line of scrimmage (see Figure \ref{fig:formation} (bottom) for example defensive formation with a NT). Typically, teams will employ either one NT or two DTs on the interior with DEs along either side of the defensive line. In addition to defensive lineman, other positions may attempt to rush the QB on blitzing plays such as outside linebackers (OLBs), interior linebackers (ILBs), and potentially members of the secondary (cornerback, free safety, strong safety) who typically play pass coverage.

@@

Moreover, the sum of sacks, hits, and hurries is often defined as pressures.
This is better than sacks to some extent, but suffers from problems of subjectivity (e.g., whether there is an actual hurry or not).
Pro football focus (PFF) comes up with a metric called pass-rush productivity (PRP), which is a minor modification from the aforementioned pressures metric (see Table \ref{tab:metrics}).
In particular, PRP gives twice as much weight to a sack relative to hurries and hits, which is a small upgrade to pressures.
However, the choice of weights is ad-hoc and still only indicates binary outcomes, similar to the shortcomings of previous metrics.

@@

More recently proposed metrics such as time in pocket (TIP), NGS get off [@Hermsmeyer2021nfl], and pass rush win rate (PRWR) [@Burke2018Created] are substantial improvements over the less sophisticated counting statistics, but nevertheless are still imperfect.
TIP refers to how long a QB can operate within the protected space behind the offensive line, known as the pocket.
However, this measure is highly context-dependent, as it can be influenced by a number of factors such as the defensive scheme or type of passing route.
NGS get off is an aggregated statistic, illustrating how quickly a defender can get past the line of scrimmage after the snap on average.
PRWR is created using player tracking data, which is at a more granular level than previous measures.
It demonstrates whether a pass rusher is able to beat their blocking matchup before a fixed time from the snap (2.5 seconds as chosen by ESPN).
This depends on the rather arbitrary time threshold used to define a pass rush win.
In addition, once a cutoff is chosen, PRWR converts continuous data to a win-loss indicator, becoming dichotomous like most of the metrics discussed above.

@@

The peer-reviewed literature on measuring the performance of football linemen (either offensive or defensive) is scant.
@AlamarGould2008 find an association between pass completion rate and successful pass blocking by offensive linemen.
The data for this study are collected for the first three weeks of the 2007 NFL season, manually recording whether a lineman holds a block and the time it took for quarterback to throw the football.
@AlamarGoldner2011 later follow up by using manually-tracked data for the 2010 season to estimate lineman performance for different team-positions instead of individual defenders (e.g., Chicago Bears' left tackle, Pittsburgh Steelers' center, etc.).
The paper uses survival analysis to model time in pocket and completion percentage for quarterbacks, before proposing a measure for linemen's contribution to their team's passing in terms of yards gained.

### 2023-05-14 paper/02-data.Rmd
more work
more edits/writing

In the forthcoming analysis, we rely on the data from the 2023 NFL Big Data Bowl provided by the NFL Next Gen Stats tracking system.
The data corresponds to 8,557 passing plays across 122 games in the first eight weeks of the 2021 NFL regular season. 
For each play, we have information on the on-field location, speed, angle, direction, and orientation of each player on the field and the football at a rate of 10 Hz, along with event annotations for each frame such as ball snap, pass forward, and quarterback sack, to name a few.
For our investigation, we consider only the frames between the ball snap and when a pass forward or quarterback sack is recorded for each play.
We also remove all plays with multiple quarterbacks on the field, since we need a uniquely defined quarterback to compute our metric.
After preprocessing, there are 251,060 unique frames corresponding to moments of time from the start of the play at snap until the moment the quarterback either throws the pass or is sacked.

@@

Along with the tracking information, the 2023 BDB includes scouting data provided by Pro Football Focus (PFF). 
PFF's scouting data includes manually collected player-level information at the play level, such as the player's role (e.g., whether they are a pass rusher) and credited events (e.g., player is credited with hitting the QB on the play).
In this manuscript, we use the PFF data to identify 36,362 unique players designated in the "pass rush" role across all plays.
For context, Figure \ref{fig:pass_rushers_count} displays the distribution of the number of observed pass rushers involved in a play, ranging from 1 to 8 with 4 pass rushers as the most common value.
We also leverage the PFF data to count how many hits, hurries, and sacks each pass rusher is credited with across the span of observed data. 
Additionally, we use the PFF player roles to identify which players were labeled as blocking each pass rusher to adjust for opponent strength, as discussed in Section \ref{sec:methods}.

>>>>

In the forthcoming analysis, we rely on the data from the 2023 NFL Big Data Bowl provided by the NFL Next Gen Stats tracking system.
The data corresponds to 8,557 passing plays across 122 games in the first eight weeks of the 2021 NFL regular season. 
For each play, we have information on the on-field location, speed, angle, direction, and orientation of each player on the field and the football at a rate of 10 Hz, along with event annotations for each frame such as ball snap, pass forward, and quarterback sack, to name a few.

For our investigation, we consider only the frames between the ball snap and when a pass forward or quarterback sack is recorded for each play.
We also remove all plays with multiple quarterbacks on the field, since we need a uniquely defined quarterback to compute our metric.
After preprocessing, there are 251,060 unique frames corresponding to moments of time from the start of the play at snap until the moment the quarterback either throws the pass or is sacked.

@@

Along with the tracking information, the 2023 BDB includes scouting data provided by Pro Football Focus (PFF). 
This contains manually collected player-level information at the play level, such as the player's role (e.g., whether they are a pass rusher and pass blocker) and credited events (e.g., player is credited with hitting the QB on the play).
In this manuscript, we use the PFF data to identify 36,362 unique players designated in the "pass rush" role across all plays.
For context, Figure \ref{fig:pass_rushers_count} displays the distribution of the number of observed pass rushers involved in a play, ranging from 1 to 8 with 4 pass rushers (i.e. 4-man rush formation) as the most common value.
We also leverage the PFF scouting data to count how many hits, hurries, and sacks each pass rusher is credited with across the span of observed data. 
Additionally, we use the PFF player roles to identify the blocking matchup for each pass rusher in order to adjust for opponent strength, as discussed in Section \ref{sec:methods}.

### 2023-05-14 paper/03-methods.Rmd
more edits/writing

## Strain rate in materials science

@@

## Application to pass rush in football {#sec:app}

@@

\begin{equation}
\overline{\text{STRAIN}}_{j} = \frac{1}{T_j} \sum_{t=1}^{T_j} \widehat{\text{STRAIN}}_{j}(t).
\end{equation}
This can be helpful for player evaluation, as we determine the most effective pass-rushers based on their average STRAIN values in Section \ref{sec:strainbar}.
We also use average STRAIN to assess statistical properties of our metric in Section \ref{sec:statprop}.

## Multilevel model of play-level STRAIN

>>>>

## Motivation and Definition of STRAIN {#sec:defn}

@@

@@

\begin{equation}
\overline{\text{STRAIN}}_{j} = \frac{1}{T_j} \sum_{t=1}^{T_j} \widehat{\text{STRAIN}}_{j}(t).
\end{equation}
This can be helpful for player evaluation, as we determine the most effective pass rushers based on their average STRAIN values in Section \ref{sec:strainbar}.
We also use average STRAIN to assess different statistical properties of our metric in Section \ref{sec:statprop}.

## Multilevel Model for Play-Level STRAIN

### 2023-05-14 paper/04-results.Rmd
more edits/writing

## Example play

@@

On average, STRAIN appears to increase for the first 0.5 seconds of a play, followed by a decline in the next second.
STRAIN then increases again until around 2.5-3 seconds after the snap before trending down towards the end of the play.
In context, this reflects the actions that a pass-rusher initially moves towards the quarterback, but is then stopped by the offensive line while the quarterback drops back.
When the quarterback stops dropping, the rusher closes the gap and increases STRAIN, before slowing down later on.

@@

## Ranking the best pass-rushers {#sec:strainbar}

Since STRAIN is observed continuously for every play in our data, this allows us to aggregate across all frames played and compute the average STRAIN for NFL pass rushers over the course of the eight-week sample size, as discussed in Section \ref{sec:app}.
Based on the clearly distinct patterns for different positions as previously observed, we evaluate interior pass-rushers (NT's and DT's) separately from edge rushers (OLB's and DE's).
Tables \ref{tab:edge} and \ref{tab:interior} are leaderboards for the NFL's best edge and interior rushers (with at least 100 plays) rated by the average STRAIN across all frames for the first eight weeks of the 2021 regular season.
The tables also consist of the total number of hits, hurries, and sacks for each defender.

Our results show many true positives with no surprises.
Myles Garrett and TJ Watt are known as top-tier edge rushers, while Aaron Donald is undoubtedly the best interior defender in football.
Rashan Gary, a young rusher with huge potential, ranks first among all pass-rushers according to our metric.
Moreover, our leaderboards largely match the rankings of experts in the field.
For instance, there is considerable overlap between our lists and PFF's edge [@Monson2022nfl] and interior [@Linsey2022nfl] rusher rankings released after the 2021 season.
This ultimately lends credibility to our proposed metric STRAIN as a measure of pass-rushing effectiveness.

@@

We first investigate the predictability of STRAIN as a measure of pressure.
In particular, we look at how well our metric correlates with a simple measure of pressure rate, defined as the total hits, sacks and hurries per snap.
Figure \ref{fig:fig_cor_pressure} is a scatterplot of average STRAIN and pressure rate over the course the provided 8-game sample, which reveals a fairly strong correlation ($r =0.625$) between the quantities.
Furthermore, the average STRAIN for the first 4 weeks of the 2021 season is more predictive of the last 4 weeks' pressure rate ($r=0.322$) than the first 4 weeks' pressure rate ($r=0.097$), as illustrated in Figure \ref{fig:fig_predictability}.
We also analyze the stability of STRAIN over time by comparing the average STRAIN across all frames played in first and last four weeks of the 2021 NFL regular season.
It is apparent from Figure \ref{fig:fig_stability} that there is a strong correlation $(r = 0.854)$ for this relationship, illustrating that STRAIN is a stable quantity over the provided 8-week window.

@@

## Multilevel Model

>>>>

## Real-Game Illustration of STRAIN

@@

On average, STRAIN appears to increase for the first 0.5 seconds of a play, followed by a decline in the next second.
STRAIN then increases again until around 2.5-3 seconds after the snap before trending down towards the end of the play.
In context, this reflects the actions that a pass rusher initially moves towards the quarterback, but is then stopped by the offensive line while the quarterback drops back.
When the quarterback stops dropping, the rusher closes the gap and increases STRAIN, before slowing down later on.

@@

## Ranking the Best Pass Rushers {#sec:strainbar}

Since STRAIN is observed continuously for every play in our data, this allows us to aggregate across all frames played and compute the average STRAIN for NFL pass rushers over the course of the eight-week sample size, as discussed in Section \ref{sec:defn}.
Based on the clearly distinct patterns for different positions as previously observed, we evaluate interior pass rushers (NT's and DT's) separately from edge rushers (OLB's and DE's).
Tables \ref{tab:edge} and \ref{tab:interior} are leaderboards for the NFL's best edge and interior rushers (with at least 100 plays) rated by the average STRAIN across all frames for the first eight weeks of the 2021 regular season.
The tables also consist of the total number of hits, hurries, and sacks for each defender.

Our results show many true positives with no surprises.
Myles Garrett and TJ Watt are known as top-tier edge rushers, while Aaron Donald is undoubtedly the best interior defender in football.
Rashan Gary, a young rusher with huge potential, ranks first among all pass rushers according to our metric.
Moreover, our leaderboards largely match the rankings of experts in the field.
For instance, there is considerable overlap between our lists and PFF's edge [@Monson2022nfl] and interior [@Linsey2022nfl] rusher rankings released after the 2021 season.
This ultimately lends credibility to our proposed metric STRAIN as a measure of pass rushing effectiveness.

@@

We first investigate the predictability of STRAIN as a measure of pressure.
In particular, we look at how well our metric correlates with a simple measure of pressure rate, defined as the total hits, sacks and hurries per snap.
Figure \ref{fig:fig_cor_pressure} is a scatterplot of average STRAIN and pressure rate over the course the provided 8-game sample, which reveals a fairly strong correlation ($r =0.625$) between the quantities.
Hence, defenders with high STRAIN values also tend to generate more pressure toward the quarterback.
Furthermore, the average STRAIN for the first 4 weeks of the 2021 season is more predictive of the last 4 weeks' pressure rate ($r=0.322$) than the first 4 weeks' pressure rate ($r=0.097$), as illustrated in Figure \ref{fig:fig_predictability}.

We also analyze the stability of STRAIN over time by comparing the average STRAIN across all frames played for the first and last four weeks of the 2021 NFL regular season.
It is apparent from Figure \ref{fig:fig_stability} that there is a strong positive correlation $(r = 0.854)$ for this relationship.
This means that STRAIN is a highly stable football metric over the provided 8-week time window, and pass rushers appear to carry their STRAIN values with them from week to week.
Overall, STRAIN performs well in both explaining defensive pressure on the field and predicting future performance of pass rushers.

@@

## Multilevel Model Results
